---
title: "COPUS Analysis for publication"
author: "Tony Hui"
output: 
  html_document: 
    keep_md: yes
    self_contained: yes
    toc: yes
---

## 1 Processing the data

```{r, echo=F, messages=FALSE}
require(knitr)
opts_chunk$set(echo=F, fig.width = 10, fig.height = 7)
cluster_method <- "ward.D2"

suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(stringr))
suppressPackageStartupMessages(require(tidyr))
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(pheatmap))
```

```{r}
all_classes <- read.csv(file = "all_observations_cleaned.csv") %>% tbl_df()

chunks <- lapply(1:5, function(x) rep(x, 5)) %>% unlist() %>% c(., 5)

all_classes <- all_classes %>%
  group_by(course, instructor, semester, observation, year) %>%
  mutate(chunk = chunks[1:n()])
```

## 2 Cluster ignoring time

### 2.1 Cluster of all classes

The x-axis labels represent course-instructor-semester pairings:

Course 11, instructor A, semester 1 = 11-A-1

Fractional time was calculated as a mean of all observations

```{r}
all_total_classes <- all_classes %>%
  ungroup() %>%
  select(-chunk, -EffectSize.StudentPerformance, -year) %>%
  gather(key = measure, value = value, -(course:observation), -time, na.rm = T) %>%
  group_by(course, instructor, semester, measure) %>%
  summarize(frac = sum(value)/n()) 

all_total_classes_matrix <- all_total_classes %>%
  mutate(id = paste(course, instructor,semester, sep = "-")) %>%
  ungroup() %>%
  select(-course, -instructor, -semester) %>%
  spread(key = measure, value = frac)
```

```{r all_years_all_times_cluster}
cluster_matrix <- all_total_classes_matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- all_total_classes_matrix$id

annotation <- data.frame(Student_perf = all_classes$EffectSize.StudentPerformance %>% unique)

rownames(annotation) <- all_total_classes_matrix$id

color_scale <- colorRampPalette(c("#ffeda0", "#feb24c", "#f03b20"))(5)
color_diverging <- colorRampPalette(c("#d7191c", "#ffffbf", "#2b83ba"))(5)

ann_colors = list(
    Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
)

pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         annotation_col = annotation,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

### 2.2 Cluster of first-year classes

```{r}
course_of_interest <- c(11,12)
first_year_classes_matrix <- all_total_classes %>%
  filter(course %in% course_of_interest) %>%
  mutate(id = paste(course, instructor,semester, sep = "-")) %>%
  ungroup() %>%
  select(-course, -instructor, -semester) %>%
  spread(key = measure, value = frac)
```

```{r first_year_all_times_cluster}
matrix <- first_year_classes_matrix

cluster_matrix <- matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- matrix$id

annotation <- data.frame(Student_perf = filter(all_classes, course %in% course_of_interest)$EffectSize.StudentPerformance %>% unique)

rownames(annotation) <- matrix$id

ann_colors = list(
    Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
)

pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         annotation_col = annotation,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

## 3 Cluster, slicing time into 10-minute intervals

### 3.1 All classes, sliced, clustered

```{r}
all_total_classes_sliced <- all_classes %>%
  ungroup() %>%
  select(-EffectSize.StudentPerformance, -year) %>%
  gather(key = measure, value = value, -(course:observation), -time, -chunk, na.rm = T) %>%
  group_by(course, instructor, semester, measure, chunk) %>%
  summarize(frac = sum(value)/n()) 

all_total_classes_sliced_matrix <- all_total_classes_sliced %>%
  mutate(id = paste(course, instructor, semester, paste0("chunk",chunk), sep = "-")) %>%
  ungroup() %>%
  select(-course, -instructor, -semester, -chunk) %>%
  spread(key = measure, value = frac)
```

```{r all_years_sliced_times_cluster}
matrix <- all_total_classes_sliced_matrix

cluster_matrix <- matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- matrix$id

annotation_raw <- all_classes %>% ungroup %>% select(course, instructor, semester, year, chunk, EffectSize.StudentPerformance) %>% 
  mutate(semester = factor(semester),
         course = factor(course),
         year = factor(year)) %>% 
  select(-semester, -year) %>%
  unique() %>% as.data.frame()

rownames(annotation_raw) <- matrix$id

# ann_colors = list(
#     Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
# )

pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         annotation_col = annotation_raw,
         # annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

### 3.2 first year classes, sliced, clustered

```{r}
first_year_classes_sliced <- all_classes %>%
  filter(course %in% course_of_interest) %>%
  ungroup() %>%
  select(-EffectSize.StudentPerformance, -year) %>%
  gather(key = measure, value = value, -(course:observation), -time, -chunk, na.rm = T) %>%
  group_by(course, instructor, semester, measure, chunk) %>%
  summarize(frac = sum(value)/n()) 

first_year_classes_sliced_matrix <- first_year_classes_sliced %>%
  mutate(id = paste(course, instructor, semester, paste0("chunk",chunk), sep = "-")) %>%
  ungroup() %>%
  select(-course, -instructor, -semester, -chunk) %>%
  spread(key = measure, value = frac)
```

```{r first_year_sliced_times_cluster}
matrix <- first_year_classes_sliced_matrix

cluster_matrix <- matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- matrix$id

annotation_raw <- all_classes %>% ungroup %>% select(course, instructor, semester, year, chunk, EffectSize.StudentPerformance) %>% 
  filter(course %in% course_of_interest) %>%
  mutate(semester = factor(semester),
         course = factor(course)) %>% 
  select(-semester, -year) %>%
  unique() %>% as.data.frame()

rownames(annotation_raw) <- matrix$id

ann_colors = list(
    Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
)

pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         annotation_col = annotation_raw,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

## 4 Cluster within classes using different chunks as features

### 4.1 first year and second year classes, sliced, clustered based on classes X chunk*metric

```{r}
jr_class <- unique(all_classes$course)[1:5]

jr_classes_sliced <- all_classes %>%
  filter(course %in% jr_class) %>%
  ungroup() %>%
  select(-EffectSize.StudentPerformance, -year) %>%
  gather(key = measure, value = value, -(course:observation), -time, -chunk, na.rm = T) %>%
  group_by(course, instructor, semester, measure, chunk) %>%
  summarize(frac = sum(value)/n()) 

jr_classes_sliced_matrix_long <- jr_classes_sliced %>%
  ungroup() %>%
  mutate(id = paste(course, instructor, semester, sep = "-"),
         measure = paste(measure, paste0("chunk",chunk), sep = "-")) %>%
  select(-course, -instructor, -semester, -chunk) %>%
  spread(key = measure, value = frac)
```

```{r jr_sliced_times_cluter_rows}
matrix <- jr_classes_sliced_matrix_long

cluster_matrix <- matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- matrix$id

annotation_raw <- all_classes %>% ungroup %>% select(course, instructor, semester, year, EffectSize.StudentPerformance) %>% 
  filter(course %in% jr_class) %>%
  mutate(semester = factor(semester),
         course = factor(course)) %>% 
  select(-semester) %>%
  unique() %>% as.data.frame()

rownames(annotation_raw) <- matrix$id

annotation_raw_order <- annotation_raw[order(annotation_raw$EffectSize.StudentPerformance),]

ann_colors = list(
    Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
)

cluster_matrix <- cluster_matrix[rownames(annotation_raw_order),]

jr_class_results <- pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         annotation_col = annotation_raw,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

### 4.2 No clustering on columns

```{r}
pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         cluster_cols = FALSE,
         # cluster_rows = FALSE,
         annotation_col = annotation_raw,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

### 4.3 No clustering on rows

```{r}
pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         cluster_cols = TRUE,
         cluster_rows = FALSE,
         annotation_col = annotation_raw,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

### 4.4 first year only

```{r}
first_yr_class <- unique(all_classes$course)[1:2]

first_yr_classes_sliced <- all_classes %>%
  filter(course %in% first_yr_class) %>%
  ungroup() %>%
  select(-EffectSize.StudentPerformance, -year) %>%
  gather(key = measure, value = value, -(course:observation), -time, -chunk, na.rm = T) %>%
  group_by(course, instructor, semester, measure, chunk) %>%
  summarize(frac = sum(value)/n()) 

first_yr_classes_sliced_matrix_long <- first_yr_classes_sliced %>%
  ungroup() %>%
  mutate(id = paste(course, instructor, semester, sep = "-"),
         measure = paste(measure, paste0("chunk",chunk), sep = "-")) %>%
  select(-course, -instructor, -semester, -chunk) %>%
  spread(key = measure, value = frac)

matrix <- first_yr_classes_sliced_matrix_long

cluster_matrix <- matrix %>% select(-id) %>% data.matrix()

rownames(cluster_matrix) <- matrix$id

annotation_raw <- all_classes %>% ungroup %>% select(course, instructor, semester, year, EffectSize.StudentPerformance) %>% 
  filter(course %in% first_yr_class) %>%
  mutate(semester = factor(semester),
         course = factor(course)) %>% 
  select(-semester) %>%
  unique() %>% as.data.frame()

rownames(annotation_raw) <- matrix$id

annotation_raw_order <- annotation_raw[order(annotation_raw$EffectSize.StudentPerformance),]

ann_colors = list(
    Student_perf = colorRampPalette(c("#edf8fb", "#006d2c"))(5)
)

cluster_matrix <- cluster_matrix[rownames(annotation_raw_order),]

first_yr_class_results <- pheatmap(mat = t(cluster_matrix), 
         clustering_method = cluster_method,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         annotation_col = annotation_raw,
         annotation_colors = ann_colors,
         # scale = "row",
         color = color_scale)
```

## 5 Rule association mining

From first year classes only.

```{r}
annotation_raw <- all_classes %>% ungroup %>% select(course, instructor, semester, year, EffectSize.StudentPerformance) %>% 
  # filter(course %in% jr_class) %>%
  filter(course %in% first_yr_class) %>%
  mutate(semester = factor(semester),
         course = factor(course),
         id = paste(course, instructor, semester, sep = "-")) %>% 
  select(-semester) %>%
  unique() %>% as.data.frame()

cutoff <- quantile(annotation_raw$EffectSize.StudentPerformance)
annotation_raw_groups <- annotation_raw %>%
  mutate(cluster = as.numeric(EffectSize.StudentPerformance > cutoff[3])) %>%
  select(id, cluster)

annotation_raw %>%
  arrange(EffectSize.StudentPerformance) %>%
  mutate(cluster = as.numeric(EffectSize.StudentPerformance > cutoff[3])) %>% 
  kable("markdown")
```

```{r, fig.width=5, fig.height=8, dpi=150}
jr_class_association_mining <- first_yr_classes_sliced %>%
  ungroup() %>%
  mutate(id = paste(course, instructor, semester, sep = "-"),
         measure = paste(measure, paste0("chunk",chunk), sep = "-")) %>%
  select(-course, -instructor, -semester, -chunk) %>%
  left_join(., annotation_raw_groups)
  # spread(key = measure, value = frac)
```

See which "features" are significantly different between the two clusters

```{r}
jr_class_association_mining_pval <- jr_class_association_mining %>%
  group_by(measure) %>%
  arrange(measure) %>%
  summarize(pval = t.test(x = frac[cluster == 0], y= frac[cluster == 1])$p.value %>% round(3)) %>%
  arrange(pval) %>%
  mutate(fdr = p.adjust(pval))

jr_class_association_mining_pval %>% head %>% kable("markdown")

measures <- jr_class_association_mining_pval %>% filter(pval < 0.1) %>% .$measure
```

Visualize the chunks with pvalue < 0.1

```{r}
jr_class_association_mining %>%
  filter(measure %in% measures) %>%
  ggplot(aes(factor(cluster), frac)) +
  geom_boxplot() +
  facet_wrap(~measure, scale = "free_y", ncol=2)
```

## Heatmap over time for the two clusters

Average the of all classes that fall within cluster 1 or 2 and plot over time.

```{r}
# jr_class_heatmap_over_time <- jr_classes_sliced %>%
jr_class_heatmap_over_time <- first_yr_classes_sliced %>%
  ungroup() %>%
  mutate(id = paste(course, instructor, semester, sep = "-")) %>%
  select(-course, -instructor, -semester) %>%
  left_join(., annotation_raw_groups) %>%
  group_by(cluster, measure, chunk) %>%
  summarize(frac = mean(frac)) %>%
  ungroup() %>%
  mutate(measure = paste(cluster, measure, sep = "_"))

jr_class_heatmap_over_time_matrix <- jr_class_heatmap_over_time %>%
  select(-cluster) %>%
  spread(key = measure, value = frac) %>%
  select(-chunk) %>%
  ungroup() %>%
  data.matrix()

rownames(jr_class_heatmap_over_time_matrix) <- rep(c(1:5), 1)

jr_class_heatmap_over_time_matrix %>%
  # head(5) %>%
  t() %>%
  pheatmap(gaps_row = 8,
           cluster_cols = FALSE,
           cluster_rows = FALSE,
           # main = "Heatmap over time for cluster 1",
           clustering_method = cluster_method,
           # scale = "row",
           color = color_scale)
```

